Towards Patient-Specific Neural Field Modeling using Micro-Electrode Data

### 2011-02-12 ltx/data_paper/DataPaper.tex
edits to maximization

Having approximated $q\left(\mathbf x_t\right)$ in the E-step, by substituting  \ref{eq:Qfunctiontrace1} and \ref{eq:Qfunctiontrace2} into \ref{eq:QIntermsofJointDist} and taking the expectation of the log likelihood function for all time instants gives an approximate to the lower-bound that needs to be maximised to gain optimal parameter estimates. Therefore $\mathcal Q$-function can be written as  
\begin{align}\label{eq:Voldermont}
 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&= -(T-1)n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&-(T-1)n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\frac{2\xi}{\sigma_d^2} \mathrm{tr}\left\lbrace \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right] \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace \nonumber \\
&-\frac{\xi^2}{\sigma_d^2}\mathrm{tr} \left\lbrace\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace +\frac{2}{\sigma_d^2}\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q( \mathbf x_t)\right]\boldsymbol\theta
 \nonumber \\
&-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}  q^\top(\mathbf  x_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)\right]\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)  \right] \boldsymbol\theta
It should be noted that the second term in equation \ref{eq:Voldermont} is obtained by adding and subtracting $\mathbf C \mathbf{\hat{x}}_{t+1} \mathbf{\hat{x}}_{t+1}^\top \mathbf C^\top$ to the expectation, $\mathbf E_{\boldsymbol\Theta'}\left[ .\right] $. For simplicity we write \ref{eq:Voldermont} in a form of
\begin{align}\label{eq:VoldermontwitXivariables}
 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&=-(T-1)n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&-(T-1)n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace \Xi_{0} \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\frac{2\xi}{\sigma_d^2} \mathrm{tr}\left\lbrace \Xi_{1} \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -  \frac{\xi^2}{\sigma_d^2}\mathrm{tr} \left\lbrace\Xi_{2}\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace 
 \nonumber \\
&+\frac{2}{\sigma_d^2}\Xi_{3}\boldsymbol\theta-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \Xi_{4}\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \Xi_{5} \boldsymbol\theta
where
	\Xi_{0}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]\label{eq:defofXi0} \\
\Xi_{1}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]\label{eq:defofXi1} \\
\Xi_{2}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]\\
\Xi_{3}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q( \mathbf x_t)\right] \label{eq:Xi3} \\	
\Xi_{4}&= \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}  q^\top(\mathbf  x_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)\right] \label{eq:Xi4}\\
 \Xi_{5}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)  \right] \label{eq:Xi5}
Consider first the maximization of $\mathcal{Q}$-function with respect to $\boldsymbol\theta$, we can write
\begin{equation}\label{eq:Qpartialtheta}
 \frac{\partial \mathcal Q}{\partial \boldsymbol\theta}=\frac{2}{\sigma_d^2}\Xi_3-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top(\Xi_4^\top+\Xi_4)-\frac{2\xi}{\sigma_d^2}\Xi_5
noting that $\Xi_4$ is symmetric and positive-definite by construction, equating \ref{eq:Qpartialtheta} to the zero vector we get
\begin{equation}\label{eq:theta}
 \boldsymbol \theta=\Xi_4^{-1}\left(\Xi_3^\top-\xi\Xi_5^\top \right).
The second derivative is negative definite at this stationary point, and hence it is a global maximizer of $\mathcal{Q}$-function.
Now maximizing with respect to $\sigma_{\epsilon}^2$ we get
  \frac{\partial \mathcal Q}{\partial \sigma_{\epsilon}^2}=-(T-1)n_y\frac{1}{\sigma_{\epsilon}^2}+\frac{1}{\sigma_{\epsilon}^4}\mathrm{tr}\left\lbrace\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace
setting above to zero and solving for $\sigma_{\epsilon}^2$ we have
 \sigma_{\epsilon}^2=\frac{1}{(T-1)n_y}\sum_{t=0}^{T-1}\mathrm{tr}\left\lbrace (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top \right\rbrace
A second derivative test shows that this is indeed the maximum. Finally differentiating with respec to $\sigma_d^2$ we get
 \frac{\partial \mathcal Q}{\partial \sigma_d^2}&= -(T-1)n_x\frac{1}{\sigma_d^2}+\frac{1}{\sigma_d^4}\mathrm{tr}\left\lbrace \Xi_0 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\frac{2\xi}{\sigma_d^4} \mathrm{tr}\left\lbrace \Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace+\frac{\xi^2}{\sigma_d^4}\mathrm{tr} \left\lbrace\Xi_2\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace  \nonumber \\
&-\frac{2}{\sigma_d^4}\Xi_3\boldsymbol\theta +\frac{1}{\sigma_d^4}\boldsymbol\theta^\top \Xi_4\boldsymbol\theta+\frac{2\xi}{\sigma_d^4} \Xi_5 \boldsymbol\theta 
setting above to zero and solving for $\sigma_e^2$ we have
\sigma_d^2=\frac{1}{(T-1)n_x}\left[ \mathrm{tr}\left\lbrace \Xi_0 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-
2\xi\mathrm{tr}\left\lbrace \Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\xi^2\mathrm{tr} \left\lbrace\Xi_2\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -2\Xi_3\boldsymbol\theta+\boldsymbol\theta^\top \Xi_4\boldsymbol\theta+2\xi\Xi_5 \boldsymbol\theta\right]  
The second derivative test confirms this is a maximum.

@@

Having approximated $q\left(\mathbf x_t\right)$ in the E-step, by substituting  \ref{eq:qTaylor} into equations \ref{eq:Xi3}-\ref{eq:Xi5} an approximate to the non-linear $\Xi$-variables can be calculated 
\begin{align}
	\Xi_{0}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf P_{t+1}+\mathbf{\hat x}_{t+1}\mathbf{\hat x}_{t+1}^\top\right)\label{eq:defofXi0} \\
\Xi_{1}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \label{eq:defofXi1} \\
\Xi_{2}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right)\\
\Xi_{3}&=\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_{t+1} \tilde{\boldsymbol\Sigma}_e^{-1} \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] \\	
\Xi_{4}&=\sum_{t=0}^{T-1}\left[q^\top(\mathbf{\hat x}_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\Lambda_t(\mathbf r)\mathbf P_t \tilde{\Lambda}_t(\mathbf r)\right] \\
 \Xi_{5}&=\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf P_t \tilde{\boldsymbol\Sigma}_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] 
 \end{align}
where
\begin{align}
	 \Lambda_t(\mathbf r)&=\begin{bmatrix}\boldsymbol \lambda_t^{(1)}(\mathbf r) & \boldsymbol \lambda_t^{(2)}(\mathbf r)& \dots &\boldsymbol \lambda_t^{(i)}(\mathbf r)\end{bmatrix} \label{eq:Lambda}\\
	 \boldsymbol\lambda_t^{(i)}(\mathbf r)&=\int_{\Omega} \boldsymbol \Psi^\top(\mathbf r')\tilde{\boldsymbol\Sigma}_e^{-1}\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t)d\mathbf r'\label{eq:lambdai}\\
	\tilde{\Lambda}_t(\mathbf r)&=\begin{bmatrix}\tilde{\boldsymbol \lambda}_t^{(1)}(\mathbf r) \\ \tilde{\boldsymbol \lambda}_t^{(2)}(\mathbf r) \\ \vdots \\ \tilde{\boldsymbol \lambda}_t^{(i)}(\mathbf r)\end{bmatrix} \label{eq:Lambdatilde}\\
	 \tilde{\boldsymbol \lambda}_t^{(i)}(\mathbf r)&=\int_{\Omega} \boldsymbol \Psi(\mathbf r')\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t)d\mathbf r' \label{eq:lambdatildei}
\end{align}
Given the state estimates, $\mathbf {\hat x^{b}}_t$ the state covariance $\mathbf P_t^b$ and cross-covariance $\mathbf M_t^b$ , the $\Xi$-variables can be computed directly. In the M-step, an update $\boldsymbol\Theta'$ is chosen such that $\mathcal{Q}\left(\boldsymbol\Theta',\boldsymbol\Theta'\right)>\mathcal{Q}\left(\boldsymbol\Theta,\boldsymbol\Theta'\right)$, this will guarantee that the likelihood increases.

>>>>

Substituting  \ref{eq:Qfunctiontrace1} and \ref{eq:Qfunctiontrace2} into \ref{eq:QIntermsofJointDist} and taking the expectation of the log likelihood function for all time instants gives the lower-bound that needs to be maximised to gain optimal parameter estimates,
\begin{align}\label{eq:Voldermont}
 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&= -(T-1)n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&-(T-1)n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\frac{2\xi}{\sigma_d^2} \mathrm{tr}\left\lbrace \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right] \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace \nonumber \\
&-\frac{\xi^2}{\sigma_d^2}\mathrm{tr} \left\lbrace\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace +\frac{2}{\sigma_d^2}\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q( \mathbf x_t)\right]\boldsymbol\theta
 \nonumber \\
&-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}  q^\top(\mathbf  x_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)\right]\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)  \right] \boldsymbol\theta
\end{align}
It should be noted that the second term in equation \ref{eq:Voldermont} is obtained by adding and subtracting $\mathbf C \mathbf{\hat{x}}_{t+1} \mathbf{\hat{x}}_{t+1}^\top \mathbf C^\top$ to the expectation, $\mathbf E_{\boldsymbol\Theta'}\left[ .\right]$. By choosing
\begin{align}
	\Xi_{0}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]\label{eq:defofXi0} \\
\Xi_{1}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]\label{eq:defofXi1} \\
\Xi_{2}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]\\
\Xi_{3}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q( \mathbf x_t)\right] \label{eq:Xi3} \\	
\Xi_{4}&= \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}  q^\top(\mathbf  x_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)\right] \label{eq:Xi4}\\
 \Xi_{5}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)  \right] \label{eq:Xi5}
 \end{align}
equation \ref{eq:Voldermont} can be written as
\begin{align}\label{eq:VoldermontwitXivariables}
 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&=-(T-1)n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&-(T-1)n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace \Xi_{0} \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\frac{2\xi}{\sigma_d^2} \mathrm{tr}\left\lbrace \Xi_{1} \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -  \frac{\xi^2}{\sigma_d^2}\mathrm{tr} \left\lbrace\Xi_{2}\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace 
 \nonumber \\
&+\frac{2}{\sigma_d^2}\Xi_{3}\boldsymbol\theta-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \Xi_{4}\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \Xi_{5} \boldsymbol\theta.
\end{align}

Consider first the maximization of $\mathcal{Q}$-function with respect to $\boldsymbol\theta$, we can write
\begin{equation}\label{eq:Qpartialtheta}
 \frac{\partial \mathcal Q}{\partial \boldsymbol\theta}=\frac{2}{\sigma_d^2}\Xi_3-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top(\Xi_4^\top+\Xi_4)-\frac{2\xi}{\sigma_d^2}\Xi_5,
\end{equation}
variable $\Xi_4$ is symmetric and positive-definite by construction, equating \ref{eq:Qpartialtheta} to the zero vector yields
\begin{equation}\label{eq:theta}
 \boldsymbol \theta=\Xi_4^{-1}\left(\Xi_3^\top-\xi\Xi_5^\top \right).
\end{equation}
The second derivative is negative definite at this stationary point, and hence it is a global maximizer of $\mathcal{Q}$-function.
Now maximizing with respect to $\sigma_{\epsilon}^2$ we get
\begin{equation}
  \frac{\partial \mathcal Q}{\partial \sigma_{\epsilon}^2}=-(T-1)n_y\frac{1}{\sigma_{\epsilon}^2}+\frac{1}{\sigma_{\epsilon}^4}\mathrm{tr}\left\lbrace\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace
\end{equation}
setting above to zero and solving for $\sigma_{\epsilon}^2$ gives
\begin{equation}
 \sigma_{\epsilon}^2=\frac{1}{(T-1)n_y}\sum_{t=0}^{T-1}\mathrm{tr}\left\lbrace (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top \right\rbrace
\end{equation}
A second derivative test shows that this is indeed the maximum. Finally differentiating with respec to $\sigma_d^2$ we get
\begin{align}
 \frac{\partial \mathcal Q}{\partial \sigma_d^2}&= -(T-1)n_x\frac{1}{\sigma_d^2}+\frac{1}{\sigma_d^4}\mathrm{tr}\left\lbrace \Xi_0 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\frac{2\xi}{\sigma_d^4} \mathrm{tr}\left\lbrace \Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace+\frac{\xi^2}{\sigma_d^4}\mathrm{tr} \left\lbrace\Xi_2\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace  \nonumber \\
&-\frac{2}{\sigma_d^4}\Xi_3\boldsymbol\theta +\frac{1}{\sigma_d^4}\boldsymbol\theta^\top \Xi_4\boldsymbol\theta+\frac{2\xi}{\sigma_d^4} \Xi_5 \boldsymbol\theta 
\end{align}
setting above to zero and solving for $\sigma_e^2$ we have
\begin{align}
\sigma_d^2=\frac{1}{(T-1)n_x}\left[ \mathrm{tr}\left\lbrace \Xi_0 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-
2\xi\mathrm{tr}\left\lbrace \Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\xi^2\mathrm{tr} \left\lbrace\Xi_2\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -2\Xi_3\boldsymbol\theta+\boldsymbol\theta^\top \Xi_4\boldsymbol\theta+2\xi\Xi_5 \boldsymbol\theta\right]  
\end{align}
The second derivative test confirms this is a maximum.

@@

Having approximated $q\left(\mathbf x_t\right)$ in the E-step, by substituting  \ref{eq:qTaylor} into equations \ref{eq:Xi3}-\ref{eq:Xi5} an approximate to the non-linear $\Xi$-variables can be calculated which provides an approximate lower bound on the log-likelihood function (see supporting information), 
\begin{align}
\Xi_{3}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_{t+1} \tilde{\boldsymbol\Sigma}_e^{-1} \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] \\	
\Xi_{4}&\approx\sum_{t=0}^{T-1}\left[q^\top(\mathbf{\hat x}_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\Lambda_t(\mathbf r)\mathbf P_t \tilde{\Lambda}_t(\mathbf r)\right] \\
 \Xi_{5}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf P_t \tilde{\boldsymbol\Sigma}_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] 
 \end{align}
where
\begin{align}
	 \Lambda_t(\mathbf r)&=\begin{bmatrix}\boldsymbol \lambda_t^{(1)}(\mathbf r) & \boldsymbol \lambda_t^{(2)}(\mathbf r)& \dots &\boldsymbol \lambda_t^{(i)}(\mathbf r)\end{bmatrix} \label{eq:Lambda}\\
	 \boldsymbol\lambda_t^{(i)}(\mathbf r)&=\int_{\Omega} \boldsymbol \Psi^\top(\mathbf r')\tilde{\boldsymbol\Sigma}_e^{-1}\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t)d\mathbf r'\label{eq:lambdai}\\
	\tilde{\Lambda}_t(\mathbf r)&=\begin{bmatrix}\tilde{\boldsymbol \lambda}_t^{(1)}(\mathbf r) \\ \tilde{\boldsymbol \lambda}_t^{(2)}(\mathbf r) \\ \vdots \\ \tilde{\boldsymbol \lambda}_t^{(i)}(\mathbf r)\end{bmatrix} \label{eq:Lambdatilde}\\
	 \tilde{\boldsymbol \lambda}_t^{(i)}(\mathbf r)&=\int_{\Omega} \boldsymbol \Psi(\mathbf r')\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t)d\mathbf r' \label{eq:lambdatildei}
\end{align}
Given the state estimates, $\mathbf {\hat x}_t$ the state covariance $\mathbf P_t$ and cross-covariance $\mathbf M_t$ , the $\Xi$-variables can be computed directly. In the M-step, an update $\boldsymbol\Theta'$ is chosen such that $\mathcal{Q}\left(\boldsymbol\Theta',\boldsymbol\Theta'\right)>\mathcal{Q}\left(\boldsymbol\Theta,\boldsymbol\Theta'\right)$, this will guarantee that the likelihood increases.
